---
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r warning=FALSE, message = FALSE}
library(moments) 
library(corrplot)
library(rmarkdown); library(knitr); library(moments); 
library(scatterplot3d); library(corrplot); library(pso)
library(psych); library(GPArotation); library(lavaan)
```
## Airline Flight Delays - SEM

Hunter Dunn and Breanna Ranglall

```{r}
#Read in Files
Flights <- read.csv("C:/Users/brezy/OneDrive - University of St. Thomas/Certificates/flights with Top 50.csv")
Flights <- Flights[,c("DAY_OF_WEEK", "Top50", "SCHEDULED_TIME", "ELAPSED_TIME", "DISTANCE", "AIR_SYSTEM_DELAY", "SECURITY_DELAY", "LATE_AIRCRAFT_DELAY", "WEATHER_DELAY", "ARRIVAL_DELAY")]

#Pulling Data from just January
Flights <- Flights[1:469969,]

#Remove missing data
Flights <- Flights |>
  na.omit(Flights)

#Create covariance plot
SIGMA <- cov(Flights)
SIGMA
```

## Structural Equation Model
```{r}
EQN <- '
        TimeAndDistance =~ SCHEDULED_TIME + ELAPSED_TIME + DISTANCE
        
        ARRIVAL_DELAY ~ DAY_OF_WEEK + Top50 + AIR_SYSTEM_DELAY + SECURITY_DELAY + WEATHER_DELAY + TimeAndDistance
       '

MOD <- sem(EQN, sample.cov = SIGMA, sample.nobs = 469969)
MOD
```

The degrees of freedom is 18.


## Prominent Indicators of Included Multidimensional Factors
```{r}
parameterEstimates(MOD)[1:3,]
```
'Elapsed Time' and 'Distance' are both have a significant on 'Time and Distance' with a p-value <= 0.00, while 'Scheduled Time' has an NA p-value due to it being a fixed value. 'Scheduled Time' has the strongest effect on 'Time and Distance' with a coefficient of 1. However, 'Distance' is clos behind with a coefficient of 0.998. 'Distance' has a very high coefficient of 7.704, most likely due to it being measured in larger units and having a stronger influence on 'TimeAndDistance' compared to the time-related measurements in the model.


## Significant Predictors of Response Variables
```{r}
parameterEstimates(MOD)[4:9,]
```
'Day of Week', 'Top50', 'Air System Delay', 'Security Delay', 'Late Aircraft Delay', and 'Weather Delay' all have a significant effect on the response variable 'Arrival Delay' with a p-value <= 0.00. However, 'Weather Delay' and 'Top50' has the strongest effect with coefficients of 0.855 and -0.720 respectively. 'Security Delay' and 'Day of Week' has the weakest with coefficients of 0.364 and 0.374 respectively.


## Summarize and Interpret Model Fit
```{r}
fitMeasures(MOD)["mfi"]
```
Because the MFI of 0.776 is below the common threshold of 0.95, this would indicate that this model does not provide a good fit to the data

```{r}
fitMeasures(MOD)["nnfi"]
```
Because the NNFI of 0.891 is below the common threshold of 0.90, this would indicate that this model does not provide a good fit to the data. However, it is so close to the threshold, it could be considered a good fit.

```{r}
fitMeasures(MOD)["rmsea"]
```
Because the RMSEA of 0.168 is above the common threshold of 0.10, this would indicate that this model does not provide a good fit to the data.

```{r}
fitMeasures(MOD)["gfi"]
```
Approximately 82.4% of the variance and covariance in the original data matrix (i.e., the covariability structure) can be explained by this model.
